Implicit yaddayadda

### 2011-10-24 papers/PLDI2012/implicitization.tex
petites modifs

\usepackage{listings}
\lstset{language=C,mathescape=true,numbers=left} % basicstyle={\rm\fontfamily{ppl}\selectfont} trop reconnaissable

@@

This article makes the following contributions:
\begin{enumerate}
\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
\item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques. (\S\ref{sec:experiments})
\end{enumerate}

\section{Bases}
\subsection{Static analysis by abstract interpretation}
\label{sec:static_analysis}
Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$. The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is often referred to as the \emph{concrete domain}. An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion. In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
%
\footnote{%
Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents. This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}

We also assume a join operator $\sqcup: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcup \abstr{y})$ for all $\abstr{x},\abstr{y}$.

Classically, one considers the control-flow graph of the program, with edges labeled with concrete transition relations (e.g. $x' = x+1$ for an instruction \lstinline|x = x+1;|), and attaches an abstract element to each control point.
A concrete transition relation $\tau \subseteq X \times X$ is replaced by an abstract \emph{forward abstract transformer} $\abstr{\tau}: \abstr{X} \rightarrow \abstr{X}$, such that
\begin{equation}
\forall \abstr{x} \in \abstr{X}, x,x' \in X,~
x \in \gamma(\abstr{x}) \land (x,x') \in \tau \implies
x' \in \gamma \circ \abstr{\tau}(\abstr{x})
\label{eqn:transformer_correct}
\end{equation}
It is easy to see that if to any control point $p \in P$ we attach an abstract element $\abstr{x}_p$ such that
(i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, the program states always lie in $\gamma(\abstr{x}_p)$ at program location~$p$.

@@

In order to compare the various techniques, we compare by inclusion the obtained
abstract values at each control points in $P_R$.

\subsection{Precision of the various techniques}

\subsection{Precision of Abstract Domains}

>>>>

\usepackage{listings}
\lstset{language=C,mathescape=true,numbers=left,numbersep=0.8ex,tabsize=2} % basicstyle={\rm\fontfamily{ppl}\selectfont} trop reconnaissable

@@

This article makes the following contributions:
\begin{enumerate}
\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
% TODO: 'We implemented' pas une contribution; rephrase.
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
\item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques. (\S\ref{sec:experiments})
\end{enumerate}

\section{Bases}
\subsection{Static analysis by abstract interpretation}
\label{sec:static_analysis}
Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$. The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is the \emph{concrete domain}. An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion. In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
%
\footnote{%
Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents. This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}

We also assume a join operator $\sqcap: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcap \abstr{y})$ for all $\abstr{x},\abstr{y}$.

Classically, one considers the control-flow graph of the program, with edges labeled with concrete transition relations (e.g. $x' = x+1$ for an instruction \lstinline|x = x+1;|), and attaches an abstract element to each control point.
A concrete transition relation $\tau \subseteq X \times X$ is replaced by an abstract \emph{forward abstract transformer} $\abstr{\tau}: \abstr{X} \rightarrow \abstr{X}$, such that
\begin{equation}
\forall \abstr{x} \in \abstr{X}, x,x' \in X,~
x \in \gamma(\abstr{x}) \land (x,x') \in \tau \implies
x' \in \gamma \circ \abstr{\tau}(\abstr{x})
\label{eqn:transformer_correct}
\end{equation}
It is easy to see that if to any control point $p \in P$ we attach an abstract element $\abstr{x}_p$ such that
(i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, when the control point is $p$, the program state always lies in $\gamma(\abstr{x}_p)$.

@@
